---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---
In this research we would like to answer if there is a behavior in steps taken.
We will look at this from a total number of steps a day, 
an average daily pattern 
and a difference in day patterns on weekdays or weekends. 

## Loading and preprocessing the data
The data is from
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip
and has been included in the repository I forked from: 
http://github.com/rdpeng/RepData_PeerAssessment1.
I used commit 80edf39c3bb508fee88e3394542f967dd3fd3270 from Oct 22 2014.  

The data consists of two months of data from an anonymous individual
collected during the months of October and November, 2012 and include
the number of steps taken in 5 minute intervals each day.

### Extract the data file
The data file activity.csv will be extracted from the zip file if it does not
already exist. The following code will download the zip file if it did not yet
exist in the context. If the activity.csv file already exists this will not do
anything, assuming this is the right version of the file now.

```{r collecting data}
if (!file.exists("activity.csv")){
    if (!file.exists("activity.zip")){
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
                      "activity.zip",
                      mode="wb")
        }
    # unzip the contents of the zip to working folder
    unzip("activity.zip", exdir=".")
    }
```

### load data into a data frame
The data has a , separated format, uses " for character data and NA for null
data. The following will read all data and place this in the steps data.frame.
```{r loading data}
steps <- read.csv("activity.csv", 
                  stringsAsFactors=FALSE, 
                  na.strings=c("NA"))
```

### Process/transform the data
Clean empty values
```{r clean data}
complete <- steps[complete.cases(steps),]
```

## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in
the data set.

1. Make a histogram of the total number of steps taken each day
```{r histogram total number of steps}
#split on days, group intervals
day_steps <- split(complete$steps, 
                   complete$date)
# sum values per day
sum_day_steps <- lapply(day_steps, sum)

# create histogram
hist(as.numeric(sum_day_steps),
     breaks=16, 
     xlab="total steps a day",
     main="Total steps taken a day")
```

2. Calculate and report the **mean** and **median** total number of steps taken per day

Mean
```{r mean steps taken per day}
mean(as.numeric(sum_day_steps))
```

Median
```{r median steps taken per day}
median(as.numeric(sum_day_steps))
```

## What is the average daily activity pattern?
We need to see a time series plot to evaluate the total number of steps 
taken each day, including an average number of steps taken across all days

Make a time series plot (i.e.  type = "l" ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r plotting time series}
#split on timeslots, group days
timeslots <- split(complete$steps, 
                   complete$interval)
# average out all day values
average <- sapply(timeslots,
                  mean)
#plot the average across all days
plot(names(average), 
     average, 
     type="l", 
     xlab="5 minute time interval", 
     ylab="average steps across all days")
```

2. Which 5-minute interval, on average across all the days in the data set, contains the maximum number of steps?
5 minute interval that has the largest number of steps
```{r maximum average steps timeslot}
names(average[average == max(average)])
```

## Imputing missing values
Note that there are a number of days/intervals where there are missing
values (coded as `NA`). The presence of missing days may introduce
bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the data set (i.e. the total number of rows with `NA`s)
Total number of missing values
```{r number of rows ith missing values}
sum(is.na(steps$steps))
```

2. Devise a strategy for filling in all of the missing values in the data set. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

I'd be filling the missing data with **mean** value of the other days for the same time interval.

3. Create a new data set that is equal to the original data set but with the missing data filled in.
fill missing values with averages
```{r filling missing values with mean values}
filled <- steps
#use name propery on average set for finding the right average value
filled$steps[is.na(filled$steps)] <- average[as.character(filled$interval[is.na(filled$steps)])]
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the **mean** and **median** total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r histogram total number of steps filled}
#split on days, group intervals
day_steps_filled <- split(filled$steps, 
                          filled$date)
# sum values per day
sum_day_steps_filled <- lapply(day_steps_filled, sum)

#render the histogram
hist(as.numeric(sum_day_steps_filled),
     breaks=16,
     xlab="total steps a day", main="Total steps taken a day")
```

```{r mean steps taken per day filled}
mean(as.numeric(sum_day_steps_filled))
```

```{r median steps taken per day filled}
median(as.numeric(sum_day_steps_filled))
```

### Changes
As expected the mean stays the same, the median is a bit higher, also as 
expected as I ignored a few values before, I did not expect it to be the
same as the mean. The histogram shows a little higher frequencies, which 
was to be expected as there are more valid values now.


## Are there differences in activity patterns between weekdays and weekends?
For this part the `weekdays()` function may be of some help here. Use
the data set with the filled-in missing values for this part.

1. Create a new factor variable in the data set with two levels -- "weekday" and "weekend"
indicating whether a given date is a weekday or weekend day.
```{r ading a weekpart factor}
library(lubridate)
filled$weekpart <-factor(ifelse(weekdays(ymd(filled$date)) == "Saturday" | 
                                    weekdays(ymd(filled$date)) == "Sunday",
                                "weekend",
                                "weekday"))
```

1. Make a panel plot containing a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

I had a few issues in the below with the list type not plotting hence I opted to
separate out all weekdays and weekends and then combining them to use
lattice plot again.

```{r plot average filled in data by weekend and weekday}
 #section of days n segment
filled_weekend <- filled[filled$weekpart=="weekend",1:3]
filled_weekdays <- filled[filled$weekpart=="weekday",1:3]

 #split on timeslots, group days
filled_weekend_timeslots <- split(filled_weekend$steps, 
                                  filled_weekend$interval)
filled_weekdays_timeslots <- split(filled_weekdays$steps, 
                                   filled_weekdays$interval)

# average out all day values
filled_weekend_average <- sapply(filled_weekend_timeslots, 
                                 mean)
filled_weekdays_average <- sapply(filled_weekdays_timeslots, 
                                  mean)

# create dataframe for lattice system, 
# weird behavior in dataframe on using strings in weekpart so using integer
filled_data <- rbind(data.frame(cbind(Interval=as.numeric(names(filled_weekend_average)), 
                                      Steps=filled_weekend_average, 
                                      weekpart=1)),
                     data.frame(cbind(Interval=as.numeric(names(filled_weekdays_average)), 
                                      Steps=filled_weekdays_average, 
                                      weekpart=2)))

#add factor for weekday, use this sequence for lattice plot
filled_data$weekpart <- factor(ifelse(filled_data$weekpart==1,
                                      "weekend",
                                      "weekday"),
                               levels=c("weekday","weekend"))

#plot
library(lattice)
xyplot(Steps~Interval|weekpart, 
       data=filled_data, 
       type="l", 
       layout=c(1,2), 
       ylab="Number of steps")
```
